**Bayesian statistics for longitudinal studies in biomedical research**

### 2021-01-03 Manuscripts/Manuscript_AM_v1.Rmd
Updated manuscript and references

As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, while enabling inferences that are both adequate and consistent from a statistical perspective.
In this regard, Generalized Additive Models (GAMs) present an alternative approach to analyze longitudinal data. While not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018]. There are several advantages of GAMs over LMEMs and rm-ANOVA models: They have an extensively developed theoretical background, enable the data to dictate the trend of the model, can accommodate missing observations and do not require constant correlation between repeated measurements [@wood2017].Therefore, GAMs can provide a more suitable statistical method to analyze non-linear biomedical longitudinal data.

@@

```{r,FIGURE 1,echo=FALSE, message=FALSE, warning=FALSE, fig.cap='Simulated longitudinal data with a linear trend.'}
rm(list=ls())
set.seed(123)
library(ggplot2)
library(splines)
library(tidyverse)
library(here)

linear<-read.csv(here("data","Simulated_linear.csv"))

sim_linear <- function(dat, n = 10, sd = 5) {
    linear_vals <- dat %>%
        slice(rep(1:n(), each = n)) %>%
        group_by(Group,Day) %>%
        mutate(observation = rnorm(n, y, sd)) %>%
        ungroup()

    return(linear_vals)
}
n<-10
sd<-2
linear_vals <- sim_linear(linear, n, sd)
linear_vals[] <- lapply(linear_vals, function(observation) ifelse(observation<=0, NA, observation)) #taking out values below 0

 ggplot(linear_vals, aes(x = Day, y = observation, color = Group)) +
    geom_point() +
    ggtitle("Simulated linear time-varying responses") +
    scale_color_viridis_d(end = 0.75) +
    geom_line(aes(y = y), alpha = 0.75) +
    theme_bw()

```

```{r,quadatric behavior}
#equation y=-2*(x-3)^2

```

@@

### Challenges presented by longitudinal studies:

@@

*make plot that explains between and within correlation*

@@

In contrast to rm-ANOVA or LMEMs, GAMs use _smooth functions_  to model the relationship between the covariates and the response. This approach is more advantageous than that of a LMEM or rm-ANOVA models as it allows a non-linear behavior of the data. One possibility is to use a polynomial function, but polynomials create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].

In GAMs the smooth functions are represented using _basis functions_  over  evenly spaced ranges of the covariates known as _knots_. The _basis function_ used is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together[@wood2017;@simpson2018]. Cubic splines have a long history in their use to solve non-parametric statistical problems and are the default choice to fit GAMs as they are the simplest option to obtain visual smoothness [@wegman1983].

This model is similar in its construction to that of \ref{eq:ANOVA}, but the key difference relies on the fact that instead of a linear combination of terms, a function is used to model the response. This allows

>>>>

As  the “frequentist” rm-ANOVA and the more advanced LMEM approach are both limited in the analysis of non-linear longitudinal information, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the information (and not an assumed trend) to determine the fit of the model, while enabling inferences that are both adequate and consistent from a statistical perspective.
In this regard, Generalized Additive Models (GAMs) present an alternative approach to analyze longitudinal data. Although not commonly used in the biomedical community, these non-parametric models have been used to analyze temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and political science [@beck1998] . There are several advantages of GAMs over LMEMs and rm-ANOVA models: They have an extensively developed theoretical background, enable the data to dictate the trend of the model, can accommodate missing observations and do not require constant correlation between repeated measurements [@wood2017].Therefore, GAMs can provide a more suitable statistical method to analyze non-linear biomedical longitudinal data.

@@

@@

### Challenges presented by longitudinal studies

@@

@@

In contrast to rm-ANOVA or LMEMs, GAMs use _smooth functions_  to model the relationship between the covariates and the response. This approach is more advantageous than that of a LMEMs or rm-ANOVA as it does not restrict the model to a linear relationship. One possible function for $f(x_t)$  is a polynomial, but a major limitation is that they create a "global" fit as they assume that the same relationship exists everywhere, which can cause problems with the fit [@beck1998].

To overcome this limitation, the smooth functions in GAMs  are represented using _basis functions_  over  evenly spaced ranges of the covariates known as _knots_. The _basis function_ used is a cubic spline, which is a smooth curve constructed from cubic polynomials joined together[@wood2017;@simpson2018]. Cubic splines have a long history in their use to solve non-parametric statistical problems and are the default choice to fit GAMs as they are the simplest option to obtain visual smoothness [@wegman1983].
